---
title: "DE Analysis for Tuberculosis"
author: "Gabe Bronk"
date: "2024-09-17"
output: html_document
---

## This script is the pipeline for differential expression analysis for tuberculosis.

## The script requires the user to input parameter values and options of their choosing in the beginning of the script (up until the point where the script says "NO MORE INPUTS ARE NEEDED BEYOND THIS POINT") - see below. 

## Before you run this script, you need to create a folder named "OutputFiles" (which will store files created by this script) in the same folder as this script is in.

## This file, DifferentialExpressionAnalysisPipelineMainScript.Rmd, loads up and organizes the data. The actual analyses are carried out in other scripts called Step1.R, Step2.R, and Step3.R. These scripts are sourced by this main script, so when you run this main script, it will automatically run the other scripts.


##------------------------------------------------------------------------------

Description/directions for each step:

1) The first step takes a minute to run. It determines if any covariates are correlated with the outcome variable (where the outcome variable is TB disease vs. no disease, TB progressor vs. non-progressor, or Mtb infected vs. uninfected). While the Step 1 script produces nominal p-values, we recommend adjusting p-values for multiple testing (e.g. take nominal p-values from all demographic/clinical covariates and run Benjamini-Hochberg correction on them).

2) This step takes about an hour to complete on a laptop. First run this with no outlier removal (i.e. see OutlierWeightingOption = 0 below). Then run DESeq2 again, this time doing outlier removal (i.e. see OutlierWeightingOption = 1 below. Also see Pre-Step 2 and Step 2 below to see how the outliers are handled).

3) This step takes a minute to complete. This step makes tables of the DE RNAs found when excluding outliers and the DE RNAs found when not excluding outliers and compares them. As discussed in our publication, we only consider RNAs to be differentially expressed if they are found to be differentially expressed in both cases. 



##------------------------------------------------------------------------------

## First set the file path:

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, warn=-1)

## Write your directory name here:
MyDirectory = 'c:/Users/gbron/Dropbox/DropboxFolders/AKlengelLab/R01Project/InfectiousDisease/ScriptsForDEPipelineSVA'

knitr::opts_knit$set(root.dir = MyDirectory)

```

## Input your selections here (i.e. which options you want to use):

```{r}
library(DESeq2)
library(readxl)
library(tictoc)
library(clusterProfiler)


AnalysisToRun = 1
## 1 is for TB disease vs. everyone else.
## 2 is for TB progressors vs. Mtb-infected non-progressors.
## 3 is for Mtb-infected vs. uninfected.
## By selecting one of these analyses, it triggers the rest of the script to use the appropriate data
## and covariates for the given analysis. If modifying this script to analyze different data, make sure ## to change the data and covariates (see the rest of this script).
## *Note that in this script, I number these analyses differently from the Comparison Numbers in the publication Table 1B.


if (AnalysisToRun == 1) {
  StringForOutputFileName = 'TBvsNoDisease'
} else if (AnalysisToRun == 2) {
  StringForOutputFileName = 'TBvsNonProgressors'
} else if (AnalysisToRun == 3) {
  StringForOutputFileName = 'InfectedvsUninfected'
}
##------------------------------------------------------------------------------

SectionToRun = 1
## This refers to which step of the script you want to run. Pick 1, 2, or 3.
## These are described above in "Description/directions for each step."

PadjThreshold = 0.10
## This is the threshold below which an adjusted p-value is considered statistically significant.

PrefilterOption = 1
## Use Option 0 to do nothing.
## Use Option 1 to exclude any genes for which there aren't at least 5 samples containing 10 (normalized)
## counts. This makes DESeq2 run a little faster. (Note that this step is not necessary because in Step 3, we ## also implement a filtering where there must be at least 44 samples containing 10 counts).

OutlierWeightingOption = 1
## Use Option 0 to do nothing.
## Use Option 1 to weight outliers with an infinitesimal weight, so that outliers are removed from coefficient and fold change estimation. By outliers, we mean that for a given gene, the outliers are samples with an extreme expression, defined by a Cook's distance cutoff. If using Option 1, you first have to run DE analysis with Option 0 in order to determine Cook's distances. 

## Choose the folder in which your output files will be stored (CREATE THIS FOLDER FIRST):
Folder = StringForOutputFileName 

## These are the names of the output files that will be saved - change as needed:
if (OutlierWeightingOption == 0) {
  FileNameStep2 = paste(Folder, '/', StringForOutputFileName, 'OutputStep2.RData', sep = "")
} else { 
  FileNameStep2 = paste(Folder, '/', StringForOutputFileName, 'OutputStep2OutliersDownweighted.RData', sep = "")
}
FileNameStep3Main = paste(Folder, '/', StringForOutputFileName, 'TableOfDERNAs', '.csv', sep = "")
FileNameStep3Supplement = paste(Folder, '/', StringForOutputFileName, 'TableOfDEResultsOutliersPresent', '.csv', sep = "")
FileNameStep3SupplementOutliersRemoved = paste(Folder, '/', StringForOutputFileName, 'TableOfDEResultsOutliersRemoved', '.csv', sep = "")




## Write the name of the folder containing data used by the script:
PathForDataToLoad = 'c:/Users/gbron/Data'

## Write technical variables here:
TechVars = c('SequencingBatch', 'ExonicRate')

```



# Loading the Data:

```{r}
setwd(PathForDataToLoad)

## The count matrix:
MyData = read.csv("RawCountMatrix.csv", header = FALSE)

## This contains covariate information:
GeneralDataOfFamilies = as.data.frame(read_excel('CovariateSpreadsheet1.xlsx'))

## These two files I only use because they tell me which shortpids correspond to which barcodes. 
## Shortpids are the IDs for each mother/child pair. Barcodes are the IDs for each sample (i.e. an individual's blood draw at a particular time point):
GeneralDataR01 = as.data.frame(read_excel('CovariateSpreadsheet2.xlsx'))
GeneralDataU01 = as.data.frame(read_excel('CovariateSpreadsheet3.xlsx'))

## This contains the sequencing, library, and extraction batch info:
BatchDF = read.csv('Batches.csv')   
## Reformatting BatchDF:
rownames(BatchDF) = BatchDF[,1]
BatchDF = BatchDF[,-1]

## This contains the information for RNA Integrity Numbers (RIN):
FinalRINDF = read.csv("RNAIntegrityNumbers.csv", header = TRUE)
## This contains other technical variables related to RNA sequencing:
TechnicalMatrix = as.data.frame(read_excel('TechnicalCovariatesFromMultiQC.xlsx'))

## This loads SmokingAndHouseDensityDataRTI, the matrix with information on household density and the composite smoking exposure score:
load('SmokingAndHouseDensityForBirthRNAKids.RData')
## RTI Stands for RNA Timepoint to Investigate (i.e. the time of blood extraction), which means we are lookings specifically at the birth timepoint.

##------------------------------------------------------------------------------
## Now we load the barcodes for the individuals. The barcode is the unique identifier for an individual at a particular time point. This loads these barcodes at the birth timepoint:
load('BarcodesOfOutcomeGroups.RData')
## The arrays loaded are:
## BarcodesOfTB refers to kids who get TB disease.
## BarcodesOfAINoIsoniazid refers to kids who become Asymptomatic Infected (i.e. non-progressors) with Mtb and did not receive prophylaxis.
## BarcodesOfInfectedOrTB refers to kids who become asymptomatic infected with Mtb or contracted TB disease (includes kids who received prophylaxis or did not). 
## BarcodesOfNonDisease refers to kids who did not contract TB (this includes kids who are asymptomatic infected who did not receive prophylaxis, and uninfected kids).
## BarcodesOfNonInfected refers to kids who never get infected (by never, I mean never within the time frame of our study). 

```


## Select what will be the topic of your analysis:

```{r}

## For BarcodesOfTetsKids and BarcodesOfControlKids, pick from the arrays listed above which ones you will be comparing if differential expression analysis:
if (AnalysisToRun == 1) {
  BarcodesOfTestKids = BarcodesOfTB 
  BarcodesOfControlKids = BarcodesOfNonDisease
} else if (AnalysisToRun == 2) {
  BarcodesOfTestKids = BarcodesOfTB 
  BarcodesOfControlKids = BarcodesOfAINoIsoniazid
} else if (AnalysisToRun == 3) {
  BarcodesOfTestKids = BarcodesOfInfectedOrTB 
  BarcodesOfControlKids = BarcodesOfNonInfected
}
```


## Set the covariates to use based on the findings of Step 1 (i.e. first run Step 1 below and then 
## fill in this section):

```{r}

if (AnalysisToRun == 1) {
                   
  CovariateArray = c('Prenatal_smoking_composite',
                     'child_hiv_birth',
                     'Food_insecurity_ANC'
                     )

}


if (AnalysisToRun == 2) {

   CovariateArray = c('Prenatal_smoking_composite',
                      'child_hiv_birth',
                      'Food_insecurity_ANC')
}


if (AnalysisToRun == 3) {
  
  CovariateArray = c(
    'Prenatal_smoking_composite',
    'birth_weight_standardized',
    'household_income_binary',
    'number_in_household_standardized',
    'sex',
    'child_hiv_birth',
    'MaternalDepression',
    'Prenatal_alcohol_composite'
  )
}
```




##==============================================================================
##==============================================================================
## NO MORE INPUTS ARE NEEDED BELOW THIS POINT 
## (i.e. Below here, you only need to edit the script if you are using 
## data different from that used in our publication 
## - see sections labeled "Edit this section if you are using data different 
## from that used in our publication")
##==============================================================================
##==============================================================================


```{r}
setwd(MyDirectory)

NumberOfTestKids = length(BarcodesOfTestKids)
NumberOfControlKids = length(BarcodesOfControlKids)
NumberOfKidsOfInterest = NumberOfTestKids + NumberOfControlKids
BarcodesOfInterest = c(BarcodesOfTestKids,BarcodesOfControlKids)

## This is used by DESeq2 when running differential expression analysis for a binary variable (e.g. TB progressor vs. non-progressor):
ContrastVariables = c("CaseOrControl",'B','A') 
## These are the groups that will be compared in DESeq2. 
## Because A is written last, A is the reference, so if we see an increase in 
## expression, it means that B is expressed more than A.
MyReferenceLevel = 'A'

```


## This section reformats the data frames:
Edit this section if you are using data different from that used in our publication.

```{r}
##------------------------------------------------------------------------------
## First we reformat the data frames containing covariate information:
##------------------------------------------------------------------------------

R01ShortpidsAndBarcodes = GeneralDataR01[,c('shortpid','Barcode')] 
U01ShortpidsAndBarcodes = GeneralDataU01[,c('shortpid','Barcode')]
R01AndU01ShortpidsAndBarcodes = rbind(R01ShortpidsAndBarcodes,U01ShortpidsAndBarcodes)

GeneralDataOfFamiliesNecessaryCovariates = GeneralDataOfFamilies[,c('shortpid', 'sex', 'R01_maternal_phenotype_breakdown', 'child_hiv_birth',	'ethnicity',	'education',	'household_income',	'work',	'Prenatal_alcohol_composite',	'smokecat_anc',	'gestation_delivery',	'preterm',	'birth_weight', 'Food_insecurity_ANC', 'Food_insecurity_14wk')]
rownames(GeneralDataOfFamiliesNecessaryCovariates) = GeneralDataOfFamiliesNecessaryCovariates[,'shortpid']

GeneralData = GeneralDataOfFamiliesNecessaryCovariates[R01AndU01ShortpidsAndBarcodes$shortpid,]
rownames(GeneralData) = R01AndU01ShortpidsAndBarcodes$Barcode
GeneralData$Barcode = R01AndU01ShortpidsAndBarcodes$Barcode

## Making Depression and PTSD be separate covariates (rather than being grouped together as "PTSD AND Depression"):
DepressionIndices = which(GeneralData$R01_maternal_phenotype_breakdown == 'Depression only')
PTSDIndices = which(GeneralData$R01_maternal_phenotype_breakdown == 'PTSD only')
PTSDAndDepressionIndices = which(GeneralData$R01_maternal_phenotype_breakdown == 'PTSD and Depression')
GeneralData[,'MaternalDepression'] = 0
GeneralData[,'MaternalPTSD'] = 0
GeneralData[DepressionIndices,'MaternalDepression'] = 1
GeneralData[PTSDAndDepressionIndices,'MaternalDepression'] = 1
GeneralData[PTSDIndices,'MaternalPTSD'] = 1
GeneralData[PTSDAndDepressionIndices,'MaternalPTSD'] = 1

##------------------------------------------------------------------------------
## Now we reformat the count matrix:
##------------------------------------------------------------------------------

## Reformatting the count matrix:
MyDataWithHeaders = MyData[-1,]
colnames(MyDataWithHeaders) = MyData[1,]
RawCountsMatrix = MyDataWithHeaders[,-1]
rownames(RawCountsMatrix) = MyDataWithHeaders[,1]

IndicesOfExpressedRNAs = which(rowSums(RawCountsMatrix)>0)

## This section arranges the Counts Matrix columns in ascending order of Barcodes:
Barcodes = colnames(RawCountsMatrix)
OrderedBarcodes = Barcodes[order(Barcodes)] 
OrderedCountsMatrix = RawCountsMatrix[IndicesOfExpressedRNAs, OrderedBarcodes]

AllBarcodes = TechnicalMatrix$Barcode
NumberOfSamples = length(AllBarcodes)

```



## Making arrays of barcodes of the individuals, so that they are ordered in ways that are convenient for the rest of the script:
Edit this section if you are using data different from that used in our publication.

```{r}

LogicalsOfNeitherPeople = AllBarcodes %in% c(BarcodesOfTestKids,BarcodesOfControlKids)
IndicesOfNeitherPeople = which(LogicalsOfNeitherPeople == FALSE)
BarcodesOfNeitherPeople = AllBarcodes[IndicesOfNeitherPeople]

CurrentBarcodesTCN = c(BarcodesOfTestKids,BarcodesOfControlKids,BarcodesOfNeitherPeople)
## TCN stands for Test, Control, Neither, since that's the order of the barcodes in this array.




AllNewbornBarcodes = c(BarcodesOfInfectedOrTB, BarcodesOfNonInfected)

IndicesOfIsoniazid = which((BarcodesOfInfectedOrTB %in% c(BarcodesOfTB, BarcodesOfAINoIsoniazid))==FALSE)
BarcodesOfIsoniazid = BarcodesOfInfectedOrTB[IndicesOfIsoniazid]

LogicalsOfOtherNewborns = 
AllNewbornBarcodes %in% c(BarcodesOfTestKids,BarcodesOfControlKids)
IndicesOfOtherNewborns = which(LogicalsOfOtherNewborns == FALSE)
BarcodesOfOtherNewborns = AllNewbornBarcodes[IndicesOfOtherNewborns]

IndicesOfOtherNewbornsNotIsoniazid = which((BarcodesOfOtherNewborns %in% BarcodesOfIsoniazid)==FALSE) 
BarcodesOfOtherNewbornsNotIsoniazid = BarcodesOfOtherNewborns[IndicesOfOtherNewbornsNotIsoniazid]

BarcodesOfBirthKids = 
c(BarcodesOfTestKids,BarcodesOfControlKids,BarcodesOfOtherNewbornsNotIsoniazid)

## Note: BarcodesOfBirthKids is the kids that will be included in the DE analysis. For AnalysisToRun = 3, this consists of all cord blood samples that we measured (376 total). For AnalysisToRun = 1 or 2, it is all the cord blood samples except the 23 kids who received isoniazid (That's because in Analyses 1 or 2, we don't assume that all the kids who received isoniazid are part of the same group (since some may have been TB progressors and some may have been non-progressors if they hadn't received isoniazid). Because we make no assumptions about these 23 kids being part of the same group and therefore cannot assign a CaseOrControl factor level to them, we have to exclude them from the count matrix in Analyses 1 and 2). 

```


## Editing the RIN Data and Technical Matrix of technical variables: 
Edit this section if you are using data different from that used in our publication.

```{r}

##------------------------------------------------------------------------------
## This section centers the RIN values because otherwise R gives me the following warning:
## The design formula contains one or more numeric variables that have mean or
## standard deviation larger than 5 (an arbitrary threshold to trigger this message).
## Including numeric variables with large mean can induce collinearity with the intercept.
## Users should center and scale numeric variables in the design to improve GLM convergence.

rownames(FinalRINDF) = FinalRINDF$Barcode
FinalRINDF$RIN = as.numeric(FinalRINDF$RIN)
MeanRIN = mean(FinalRINDF$RIN)
CenteredRINDF = FinalRINDF
CenteredRINDF$RIN = FinalRINDF$RIN - MeanRIN

##------------------------------------------------------------------------------
## Now we center and scale every Technical Variable:

rownames(TechnicalMatrix) = TechnicalMatrix$Barcode
## Column 20 is removed because I don't believe the Average Read Lengths (since all the U01 samples
## have the same average read length)
TechnicalMatrixReformatted = TechnicalMatrix[,-c(1,20)]
FinalTechnicalMatrix = TechnicalMatrixReformatted

NumberOfTechnicalVariables = dim(TechnicalMatrixReformatted)[2]
## CN stands for Column Number:
for (CN in 1:NumberOfTechnicalVariables){
  ColumnMean = mean(TechnicalMatrixReformatted[,CN])
  ColumnStdDev = sd(TechnicalMatrixReformatted[,CN])
  FinalTechnicalMatrix[,CN] = ( TechnicalMatrixReformatted[,CN] - ColumnMean )/ColumnStdDev 
}  

TypeOfBatch = 'SequencingBatch' ## This indicates that we will correct the count matrix for sequencing batch.

```


## Creating the Metadata table: 
Edit this section if you are using data different from that used in our publication.

```{r} 

##==============================================================================
## Creating the Metadata Table:
##==============================================================================

## First we include the covariates:
Metadata = data.frame('sex' = as.factor(GeneralData[CurrentBarcodesTCN,'sex']))
rownames(Metadata) = CurrentBarcodesTCN
Metadata[,'child_hiv_birth'] = as.factor(GeneralData[CurrentBarcodesTCN,'child_hiv_birth'])
Metadata[,'ethnicity'] = as.factor(GeneralData[CurrentBarcodesTCN,'ethnicity'])
Metadata[,'education'] = as.factor(GeneralData[CurrentBarcodesTCN,'education'])
Metadata[,'household_income'] = as.factor(GeneralData[CurrentBarcodesTCN,'household_income'])
Metadata[,'work'] = as.factor(GeneralData[CurrentBarcodesTCN,'work'])
Metadata[,'Prenatal_alcohol_composite'] = as.factor(GeneralData[CurrentBarcodesTCN,'Prenatal_alcohol_composite'])
Metadata[,'smokecat_anc'] = as.character(GeneralData[CurrentBarcodesTCN,'smokecat_anc'])
Metadata[,'preterm'] = as.factor(GeneralData[CurrentBarcodesTCN,'preterm'])
Metadata[,'Food_insecurity_ANC'] = as.factor(GeneralData[CurrentBarcodesTCN,'Food_insecurity_ANC'])
Metadata[,'Food_insecurity_14wk'] = as.factor(GeneralData[CurrentBarcodesTCN,'Food_insecurity_14wk'])
Metadata[,'MaternalDepression'] = as.factor(GeneralData[CurrentBarcodesTCN,'MaternalDepression'])
Metadata[,'MaternalPTSD'] = as.factor(GeneralData[CurrentBarcodesTCN,'MaternalPTSD'])
Metadata[,'gestation_delivery'] = GeneralData[CurrentBarcodesTCN,'gestation_delivery']
Metadata[,'birth_weight'] = GeneralData[CurrentBarcodesTCN,'birth_weight']
Metadata[,'Prenatal_smoking_composite'] = 
  as.factor(SmokingAndHouseDensityDataRTI[CurrentBarcodesTCN,'Prenatal_smoking_composite'])
Metadata[,'number_in_household'] = 
  as.numeric(SmokingAndHouseDensityDataRTI[CurrentBarcodesTCN,'number_in_household'])
##------------------------------------------------------------------------------

## This assigns the letter 'B' to individuals in the case group and the letter 'A' to individuals in the control group:
ArrayOfCaseOrControl = c( rep('B',length(BarcodesOfTestKids)), rep('A',length(BarcodesOfControlKids)), rep('C',length(BarcodesOfNeitherPeople)) )

Metadata[,'CaseOrControl'] = as.factor(ArrayOfCaseOrControl)
##------------------------------------------------------------------------------

## Now we add batch information and technical variables to the Metadata table:
Metadata[,'LibraryBatch'] = as.factor(BatchDF[CurrentBarcodesTCN,'Library'])
Metadata[,'SequencingBatch'] = as.factor(BatchDF[CurrentBarcodesTCN,'Sequencing'])
Metadata[,'RIN'] = CenteredRINDF[CurrentBarcodesTCN,'RIN']
NumberOfColumnsBeforeTechnicalVariables = dim(Metadata)[2]
Metadata = cbind(Metadata, FinalTechnicalMatrix[CurrentBarcodesTCN,])


##===============================================================================
## Standardizing variables so that DESeq2 doesn't get mad (because DESeq2 will spit out a warning if you don't standardize these, as discussed in the previous section about standardizing RIN):
##==============================================================================

## Standardizing birth weight:

BirthWeightArray = as.numeric(Metadata[,'birth_weight'])
MeanBirthWeight = mean(BirthWeightArray, na.rm=TRUE)
CenteredBirthWeightArray = BirthWeightArray - MeanBirthWeight

StdBirthWeight = sd(CenteredBirthWeightArray, na.rm=TRUE)
ScaledBirthWeightArray = CenteredBirthWeightArray/StdBirthWeight
Metadata[,'birth_weight_standardized'] = ScaledBirthWeightArray

## Standardizing number_in_household:

DensityArray = as.numeric(Metadata[,'number_in_household'])
MeanDensity = mean(DensityArray, na.rm=TRUE)
CenteredDensityArray = DensityArray - MeanDensity
StdDensity = sd(CenteredDensityArray, na.rm=TRUE)
ScaledDensityArray = CenteredDensityArray/StdDensity
Metadata[,'number_in_household_standardized'] = ScaledDensityArray




## I make a binary household income variable here, which I call household_income_binary:
Metadata[,'household_income_binary'] = as.character(Metadata[,'household_income'])
## I group the high income people with the middle income people:
RichIndices = which(Metadata$household_income == '>R5000/m')
Metadata[RichIndices, 'household_income_binary'] = 'middle'
MiddleIndices = which(Metadata$household_income == 'R1000-5000/m')
Metadata[MiddleIndices, 'household_income_binary'] = 'middle'
PoorIndices = which(Metadata$household_income == '<R1000/m')
Metadata[PoorIndices, 'household_income_binary'] = 'poor'
Metadata[,'household_income_binary'] = as.factor(Metadata[,'household_income_binary'])

## Here I have the ordinal household income variable, and I change the levels to the numbers
## 1, 2, and 3 rather than the strings that it had before
## (i.e. changing household income to ordinal numbers):
Metadata[, 'household_income'] = as.character(Metadata[, 'household_income'])
Metadata[RichIndices, 'household_income'] = '3'
Metadata[MiddleIndices, 'household_income'] = '2'
Metadata[PoorIndices, 'household_income'] = '1'
Metadata[,'household_income'] = as.numeric(Metadata[,'household_income'])



## Here I have the ordinal education variable, and I change the levels to the numbers
## 1, 2, 3, and 4 rather than the strings that it had before
## (i.e. changing education to ordinal numbers):
Metadata[, 'education'] = as.character(Metadata[, 'education'])
FirstIndices = which(Metadata$education == 'primary')
Metadata[FirstIndices, 'education'] = '1'
SecondIndices = which(Metadata$education == 'some secondary')
Metadata[SecondIndices, 'education'] = '2'
ThirdIndices = which(Metadata$education == 'completed secondary')
Metadata[ThirdIndices, 'education'] = '3'
FourthIndices = which(Metadata$education == 'any tertiary')
Metadata[FourthIndices, 'education'] = '4'
Metadata[,'education'] = as.numeric(Metadata[,'education'])

##------------------------------------------------------------------------------

## Adding underscores so DESeq2 doesn't give me warnings:

Metadata[,'Food_insecurity_ANC'] = as.character(Metadata[,'Food_insecurity_ANC'])
FirstIndices = which(Metadata[,'Food_insecurity_ANC'] == 'Percieved food secure')
Metadata[FirstIndices,'Food_insecurity_ANC'] = 'Percieved_food_secure'
SecondIndices = which(Metadata[,'Food_insecurity_ANC'] == 'Percieved food insecure')
Metadata[SecondIndices,'Food_insecurity_ANC'] = 'Percieved_food_insecure'
Metadata[,'Food_insecurity_ANC'] = as.factor(Metadata[,'Food_insecurity_ANC'])

Metadata[,'child_hiv_birth'] = as.character(Metadata[,'child_hiv_birth'])
FirstIndices = which(Metadata[,'child_hiv_birth'] == 'HIV exposed uninfected')
Metadata[FirstIndices,'child_hiv_birth'] = 'HIV_exposed_uninfected'
SecondIndices = which(Metadata[,'child_hiv_birth'] == 'HIV unexposed')
Metadata[SecondIndices,'child_hiv_birth'] = 'HIV_unexposed'
Metadata[,'child_hiv_birth'] = as.factor(Metadata[,'child_hiv_birth'])

##------------------------------------------------------------------------------

CurrentCountsMatrix = OrderedCountsMatrix[,CurrentBarcodesTCN]

MetadataColumnNames = colnames(Metadata)

## Technical variables are columns 60 to 76 of Metadata, but I will remove a few of them because they are very correlated with other technical variables (i.e. Pearson correlation > 0.8).
## Hence, here are the indices of the technical variables that I will consider:
TechnicalVariableIndicesToUse = c(1,2,4,7,8,9,11,12,14:17)
TechnicalVariables = MetadataColumnNames[(NumberOfColumnsBeforeTechnicalVariables+TechnicalVariableIndicesToUse)]


```


##==============================================================================
## ANALYSIS STEPS:
##==============================================================================

## Step 1
In this step, we determine if any covariates are correlated with the outcome (where the outcome is TB progressor vs non progressor, or TB disease vs no disease, or TB infection or disease vs non-infected).

```{r}
if (SectionToRun == 1) {

  source('Step1.R')

}
```


## Pre-Step 2:
## Preparation for outlier removal in DE analysis:

```{r}

  if (OutlierWeightingOption == 1) {
    
    FileNameToLoadCooks = paste(Folder, '/', StringForOutputFileName, 'OutputStep2.RData', sep = "")
    load(FileNameToLoadCooks)
    
    ##---------------------------------------
    ## Calculating cooks cut off value:
    CooksMatrix <- assays(dds)[["cooks"]]
    
    pm <- dim(CoefficientMatrix)[2]      # number of coefficients
    mm <- dim(CooksMatrix)[2]      # total sample size

    # Compute the 0.99 quantile of F(p, m-p)
    quantile_value <- qf(0.99, df1 = pm, df2 = mm-pm)
    print(quantile_value)
    ##---------------------------------------
   
    ResultsDFRENoWeighting = ResultsDFRE
    
  }
 
```


## Step 2:
## Differential Expression Analysis

```{r}
if (SectionToRun == 2) { 
  
  source('Step2.R')

}
```




## Step 3:
## Making DE results tables for the paper:

```{r}
if (SectionToRun == 3) {
  
load(FileNameStep2)

source('Step3.R')


}
```




